Hippocampal replay of experience at slow speeds

### 2021-04-27 Article.tex

% Figure 3-Supplemental 1
\figsupp[More examples of SWRs that would not be well-characterized by using the standard decoder]{A-I. More examples of SWRs that would not be well-characterized by using the standard decoder. Conventions are the same as in \FIG{3}.}{\includegraphics[width=1.0\linewidth]{figures/Figure3-supplemental1/Figure3_v4_supplemental1}}\label{figsupp:fig3sup1}

@@

% Figure 3-Supplemental 3
\figsupp[Comparison of percentage of SWRs classified on real vs. position shuffled data]{Comparison of percentage of SWRs classified (that is, a SWR containing at least one of the five classifications) on real vs. position shuffled data for two epochs from different animals. Red line represents the percent of SWR events classified in that epoch for real data. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing and position occupancy.}{\includegraphics[width=0.8\linewidth]{figures/Figure3-supplemental3/Figure3_v1-supplemental3}}\label{figsupp:fig3sup3}

@@

Having established the validity of our decoder, we next wanted to assess the overall prevalence of spatial content across SWRs. We detected SWRs using a permissive threshold (see Methods) to ensure that we include both the rare large amplitude events as well as the much more common small amplitude events. As expected, and providing further support for the validity of the model, we observe many replays that are classified as continuous throughout the entirety of the SWR, similar to those seen using the standard decoder (\FIGSUPP[2]{fig2sup2}A-F). However, we also observe many events with spatially coherent content that do not have this structure. For example, we observe trajectories that start in one direction and reverse back to the original position (\FIG{3}A, \FIGSUPP[3]{fig3sup1}B), trajectories that remain fixed in one position (\FIG{3}B, \FIGSUPP[3]{fig3sup1}G), and trajectories that jump between arms and between dynamics (\FIG{3}C, D, F, \FIGSUPP[3]{fig3sup1}F, H, I). We also observe SWRs that are spatially incoherent throughout the SWR (\FIGSUPP[3]{fig3sup1}A, D).

Using a 0.8 threshold, we are able to classify 89\% (23,071 of 25,844) of SWRs as containing at least one of the three dynamics or dynamic mixtures. To ensure that this depends on the firing of spatially tuned cells in the hippocampus, we resample position with replacement, which shuffles the relationship between spiking and position, for two recording sessions , fit the encoding model and then decode the same SWR events containing the same spikes. Only 9\% of the SWRs are classified in the shuffled datasets, a value that is significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup2}). This shows that our model will not classify dynamics when there is no spatial information present in the data.

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as SWRs containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods and \FIG{1}F). We find that 96\% (22,170 of 23,071) of classified SWRs and 86\% of all SWRs include spatially coherent structure and that this prevalence of spatially coherent structure is consistent across animals (\FIG{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We find that only 14\% (3,295 of 23,071) of classified SWRs have time points with spatially incoherent structure (\FIG{3}G).

@@

% Figure 5-Supplemental 3
\figsupp[Further quantification of spiking properties]{\textbf{(A)} Dynamics. \textbf{(B)} Ripples  }{\includegraphics[width=1.0\linewidth]{figures/Figure5-supplemental3/Figure5_v1_supplemental3}}\label{figsupp:fig5sup3}

@@

We hypothesize that this large set of events supports the ability to rapidly retrieve representations of individual places and snippets of experience. Previous findings demonstrated that SWRs with movement- or immobility-related activity engage specific patterns of spiking activity outside the hippocampus \citep{YuDistincthippocampalcorticalmemory2017, JadhavCoordinatedExcitationInhibition2016}. Thus, we would expect that the large set of events with slow dynamics would also engage specific patterns of activity outside the hippocampus, perhaps allowing animals to recall and evaluate the value of specific locations or short movement trajectories without having to generate full trajectories \citep{YuHippocampalcorticalinteraction2015}. This is consistent with our own subjective experience of memory, where we do not necessarily have to recall an entire path in detail for it to be useful, particularly if it is a well traveled route.

Our results showed that 19\% of SWRs contained continuous trajectories---meaning that portions of the SWR had spatially coherent trajectories that moved at high velocities---but only 4\% were continuous throughout the entire SWR. This, \textit{prima facie}, may seem at odds with previous results, which found between 5\% and 45\% of SWRs were "significant", meaning that they were well fit by a line compared to various shuffled versions of the dataset. Furthermore, these studies found trajectories on average move at much higher speeds (20x normal rat movement speed) than we report here. While the plurality of methods and parameters makes it hard to directly compare, we believe there are several reasons for this difference. First, our model is much more flexible because it allows the trajectory to change speed during the SWR, whereas fitting a line assumes a constant speed trajectory. So many of the events that had continuous dynamics, and would be classified as having high speed trajectories, also exhibited stationary-continuous mixture dynamics (see for example \FIG{1}). Second, the line has to be fit to the whole SWR event, meaning that it is sensitive to the estimation of the start end of the SWR event. In contrast, our model makes an estimate for each time point and only depends on the previous and future time step, which minimizes the dependence of the event on those boundaries and any possible "noise" events within the SWR.

We use a model with smaller time bins that allows for changes of speeds within SWRs, whereas methods using a linear fit use larger time bins and cannot accommodate rapid change. Studies using a linear fit typically incorporate position bins within 15 cm of the best fit line in their best fit score in order to compensate for the rigidness of this restriction, but this allows for a wide range of speeds when using a 20 ms time bin (up to 3.75 m/s). The linear fit also depends on the start and end event boundaries. Many studies explicitly excluded slow events.

>>>>

% Figure 3-Supplemental 1
\figsupp[More examples of SWRs with non-constant speed trajectories]{A-I. More examples of SWRs that would not be well-characterized by using the standard decoder. Conventions are the same as in \FIG{3}.}{\includegraphics[width=1.0\linewidth]{figures/Figure3-supplemental1/Figure3_v4_supplemental1}}\label{figsupp:fig3sup1}

@@

% Figure 3-Supplemental 3
\figsupp[Comparison of percentage of spatially coherent and incoherent SWRs on real vs. run-swapped, circularly shuffled position data]{Comparison of percentage of spatially coherent and incoherent SWRs on real vs. run-swapped, circularly shuffled position. The red line represents the percent of spatially coherent or incoherent SWRs in that epoch for actual data. The histogram represents the distribution after 50 shuffles of the run from well to well as well as the circularly shuffled position (for each tetrode). This preserves local spatial correlations in the data but breaks the global spatial relationships between the spikes and the data.}{\includegraphics[width=0.8\linewidth]{figures/Figure3-supplemental3/Figure3_v1_supplemental3}}\label{figsupp:fig3sup3}

@@

Having established the validity of our decoder in tracking the animal's position, we next wanted to assess the overall prevalence of spatial content across SWRs. We detected SWRs using a permissive threshold (see Methods) to ensure that we include both the rare large amplitude events as well as the much more common small amplitude events. As expected, and providing further support for the validity of the model, we observe many replays that are classified as continuous throughout the entirety of the SWR, similar to those seen using the standard decoder (\FIGSUPP[2]{fig2sup2}A-F). However, we also observe many events with spatially coherent content that do not have this structure. For example, we observe trajectories that start in one direction and reverse back to the original position (\FIG{3}A, \FIGSUPP[3]{fig3sup1}B), trajectories that remain fixed in one position (\FIG{3}B, \FIGSUPP[3]{fig3sup1}G), and trajectories that jump between arms and between dynamics (\FIG{3}C, D, F, \FIGSUPP[3]{fig3sup1}F, H, I). We also observe SWRs that are spatially incoherent throughout the SWR (\FIGSUPP[3]{fig3sup1}A, D).

Using a 0.8 threshold, we are able to classify 89\% (23,071 of 25,844) of SWRs as containing at least one of the three dynamics or dynamic mixtures. To ensure that this depends on the firing of spatially tuned cells in the hippocampus, we resample position with replacement, which shuffles the relationship between spiking and position, for two recording sessions, fit the encoding model and then decode the same SWR events containing the same spikes. Only 9\% of the SWRs are classified in the shuffled datasets, a value that is significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup2}). This shows that our model will not classify dynamics when there is no spatial information present in the data.

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as SWRs containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods and \FIG{1}F). We find that 96\% (22,170 of 23,071) of classified SWRs and 86\% of all SWRs include spatially coherent structure and that this prevalence of spatially coherent structure is consistent across animals (\FIG{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We find that only 14\% (3,295 of 23,071) of classified SWRs have time points with spatially incoherent structure (\FIG{3}G). To further validate this result, we performed a shuffle that randomized the order of the reward well to reward well runs and then circularly permuted the run. This, in contrast to our previous shuffle, should preserve correlation between spikes locally but destroy global spatial structure. Using this shuffle, we found that there was more shuffled spatially incoherent content compared to the real data and there was less spatially coherent content compared to the real data for two epochs (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup3}).

@@

% Figure 5-Supplemental 3
\figsupp[Further quantification of spiking and ripple properties by dynamic]{\textbf{(A)} Further quantification of the dynamics with respect to the spiking and ripple properties. \textbf{(B)} Speed and number of tetrodes with spikes for entire SWR }{\includegraphics[width=1.0\linewidth]{figures/Figure5-supplemental3/Figure5_v1_supplemental3}}\label{figsupp:fig5sup3}

@@

We hypothesize that this large set of events supports the ability to rapidly retrieve representations of individual places and snippets of experience. Previous findings demonstrated that SWRs with movement- or immobility-related activity engage specific patterns of spiking activity outside the hippocampus \citep{YuDistincthippocampalcorticalmemory2017, JadhavCoordinatedExcitationInhibition2016}. Thus, we would expect that the large set of events with slow dynamics would also engage specific patterns of activity outside the hippocampus, perhaps allowing animals to recall and evaluate the value of specific locations or short movement trajectories without having to generate full trajectories \citep{YuHippocampalcorticalinteraction2015}.

One possibility is that these events are supported by different subcircuits within the hippocampus. For example, \cite{OlivaRoleHippocampalCA22016} showed that CA2 neurons tend to lead CA1 and CA3 neurons during SWRs We did not have enough simultaneously recorded tetrodes in each area to attribute the different dynamics to particular subregions.

Our results showed that 19\% of SWRs contained continuous trajectories---meaning that portions of the SWR had spatially coherent trajectories that moved at high velocities---but only 4\% were continuous throughout the entire SWR. This, \textit{prima facie}, may seem at odds with previous results, which found between 5\% and 45\% of SWRs were "significant", meaning that they were well fit by a line compared to various shuffled versions of the dataset. Furthermore, these studies found trajectories on average move at much higher speeds (20x normal rat movement speed) than we report here. While the plurality of methods and parameters makes it hard to directly compare, we believe there are several reasons for this difference. First, our model is much more flexible because it allows the trajectory to change speed during the SWR, whereas fitting a line assumes a constant speed trajectory for the entire SWR. Many of the events that had continuous dynamics, and would be classified as having high speed trajectories, also exhibited stationary-continuous mixture dynamics (see for example \FIG{1}). Second, the line has to be fit to the whole SWR event, meaning that it is sensitive to the estimation of the start end of the SWR event. In contrast, our model makes an estimate for each time point and only depends on the previous and future time step, which minimizes the dependence of the event on those boundaries and any possible "noise" events within the SWR. This, along with our more permissive threshold for SWRs, allows us to characterize more SWRs than previous studies, rather than not analyze them.
